\chapter{Exercises - Not covered in Syllabus}
\begin{enumerate}
%abs

\item{[IEB, Nov. 2001, HG]
\begin{enumerate}
\item{Sketch two curves which will enable you to solve $|x-2| \leq |x| - 1$ graphically.}
\item{\textit{Calculate} the point(s) of intersection of the two curves, and \textit{hence}, or otherwise, write down the solution set to the inequality above.}
\end{enumerate}}

\item{[IEB, Nov. 2002, HG] Evaluate without using a calculator: $|2x-1| \: - \: |\!-\!5x|$ \quad given that $x=-1$}

\item{[IEB, Nov. 2003, HG] In the figure below, two graphs are shown:  $$f(x)=ax^2+bx+c \qquad \mathrm{ and } \qquad g(x)=|x+c|$$ $x=-1$ is the equation of the axis of symmetry of the parabola. $f$ contains the points (2; 0) and (4; -8), and $P$ is the turning point of $g$.

\begin{center}
\begin{pspicture}(0,0)(5,5)
\psgrid[gridcolor=gray]
\end{pspicture}
\end{center}

\begin{enumerate}
\item{Find the values of $a$, $b$ and $c$.}
\item{Find the length of $M\!N$ if $M\!N$ is perpendicular to the $y$-axis and $M\!N$ produced passes through the $y$-intercept of $f(x)$ and $g(x)$.}
\item{Determine the equation of the graph that results when $g(x)$ is reflected about the line $x=1$.}
\end{enumerate}}

\item{[IEB, Nov. 2003, HG]
\begin{enumerate}
\item{Sketch the graph of $f(x) = x^3 - 9x^2 + 24x - 20$, showing all intercepts with the axes and turning points.}
\item{Find the equation of the tangent to $f(x)$ at $x=4$.}
\item{Sketch the graph of $y=|f(x)|$ on a new set of axes, giving coordinates of the turning points and intercepts with the axes.}
\end{enumerate}}

\item{[IEB, Nov. 2004, HG] Solve: $18 \leq |x-3|$}

\item{[IEB, Nov. 2005, HG] Solve for $x$:\quad $25^{|1-2x|}=5^4$}

\item{[IEB, Nov. 2005, HG] If $f(x) = |4-x|$, find the value of $f^{\prime}(3)$.}
\end{enumerate}

% CHILD SECTION END 

\appendix
