\documentclass[9pt,a4paper,titlepage,twoside,openright]{extbook}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{float} % for figures to appear where you want them
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{changebar}
%\usepackage{glossaries}
\usepackage{pst-all}
\usepackage{pst-eucl}        %Jo
\usepackage{pst-poly}        %Jo
\usepackage{xcolor}
\usepackage{multicol}
\usepackage{pst-3dplot}        %Jo


%% ************* NB ************
%% The order in which pstricks packages are loaded
%% matters - so I copied the order from pst-all.sty
%% and then added the two additional packages at the end.
%% ************* End NB ************

%% ************* Packages ************
\usepackage{pst-circ}
\usepackage{pst-lens}
\usepackage{pst-optic}         %Jo
  %\usepackage{pst-labo}         %Jo
  \usepackage{subfigure}
  \usepackage{multirow}
  \usepackage{amsmath}
  \usepackage{tabularx}
  \usepackage{lscape}
  \usepackage{fancyhdr}
  \usepackage{wasysym}
  \usepackage{url}
  \usepackage{amsmath, amsthm, amsfonts, amssymb}
  \usepackage{eurosym}
  \usepackage{array}
  \usepackage{enumitem}
  \sffamily

  \usepackage{mdframed}

\input{../../DefinitionsV0-5-1}

\begin{document}


\pagenumbering{roman}
\title{The Free High School Science Texts: Textbooks for High School Students Studying the Sciences\\
\textbf{Mathematics}\\
Grade 12}
\author{FHSST Authors}
\maketitle
\cleardoublepage
\begin{quote}
Copyright 2007 ``Free High School Science Texts''\\
Permission \textbf{is} granted to copy, distribute and/or modify this document under the terms of the GNU Free Documentation License, Version 1.2 or any later version published by the Free Software Foundation; with no Invariant Sections, no Front-Cover Texts, and no Back-Cover Texts.
A copy of the license is included in the section entitled ``GNU Free Documentation License''.
\end{quote}

% This is the simple stuff about the license
\begin{center}
\vspace{1.cm}
\psshadowbox[linecolor=red,linewidth=0.25cm]{{\LARGE STOP!!!!}}
\vspace{1.cm}
{\Large \\ Did you notice the FREEDOMS we've granted you?\\}
\vspace{1.cm}
\begin{quotation}{\Large Our copyright license is \textbf{different!} It grants freedoms rather than just imposing
restrictions like all those other textbooks you probably own or use.\\}\end{quotation}
{\large 
\begin{itemize} 
\item We know people copy textbooks illegally but we would LOVE it if you copied our's - 
go ahead copy to your hearts content, \textbf{legally!}
\item Publishers revenue is generated by controlling the market, we don't want any money, 
go ahead, distribute our books far and wide - we DARE you!
\item Ever wanted to change your textbook? Of course you have! Go ahead change ours, 
make your own version, get your friends together, rip it apart and put it back together 
the way you like it. That's what we really want!
\item Copy, modify, adapt, enhance, share, critique, adore, and contextualise. Do it all, 
do it with your colleagues, your friends or alone but get involved! Together we can 
overcome the challenges our complex and diverse country presents.
\item So what is the catch? The only thing you can't do is take this book, make a few changes and then tell others that they can't do the same with your changes. It's share and share-alike and we know you'll agree that is only fair.
\item These books were written by volunteers who want to help support education, who 
want the facts to be freely available for teachers to copy, adapt and re-use. Thousands 
of hours went into making them and they are a gift to everyone in the education community.
\end{itemize}
}
\end{center}

\input{../../AuthorListV0-5}

% Stuff copied from thesis
% Dutch style of paragraph formatting, i.e. no indents. 
\setlength{\parskip}{1.3ex plus 0.2ex minus 0.2ex}
\setlength{\parindent}{0pt}


\pagestyle{fancy}
\renewcommand{\sectionmark}[1]{\markright{\thesection}{}}
\renewcommand{\footrulewidth}{0.4pt}

\cleardoublepage
\tableofcontents
\cleardoublepage
\pagenumbering{arabic}

\input{Mathematics_Gr12_Ch_1-chapterIntroductiontoBook}
\input{Mathematics_Gr12_Ch_2-Logarithms}
\input{Mathematics_Gr12_Ch_3-SequencesandSeries}
\input{Mathematics_Gr12_Ch_4-Finance}
\input{Mathematics_Gr12_Ch_5-FactorisingCubicPolynomials}
\input{Mathematics_Gr12_Ch_6-FunctionsandGraphs}
\input{Mathematics_Gr12_Ch_7-DifferentialCalculus}
\input{Mathematics_Gr12_Ch_8-LinearProgramming}
\input{Mathematics_Gr12_Ch_9-Geometry}
\input{Mathematics_Gr12_Ch_10-Trigonometry}
\input{Mathematics_Gr12_Ch_11-Statistics}
\input{Mathematics_Gr12_Ch_12-CombinationsandPermutations}
\input{Mathematics_Gr12_Ch_13-chapterGeneralExercises}
\input{Mathematics_Gr12_Ch_14-chapterExercisesNotcoveredinSyllabus}
\input{Mathematics_Gr12_Ch_15-chapterGNUFreeDocumentationLicenselabelfdl}
\end{document}
